Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-08-15 manuscript/main.tex
less wordy title
linear model in mu (not nu)

\title{Coalescent inference of mutation spectrum histories from nucleotide context-resolved sample frequency spectra}

@@

Let $C_h$ denote the event that there is a mutation with label $U\in(x,x+h)\subseteq(0,1)$ that subtends $b$ haplotypes in the sample, and let $I_k$ denote the event that this mutation arose at a time when the sample had $k$ ancestral lines.
For small $h$ we then have
\begin{align*}
\mathbbm{P}(C_h\mid T) &= \sum_{k=2}^n \mathbbm{P}(C_h, I_k\mid T) = \sum_{k=2}^n p_{n,k}(b)\mathbbm{P}(I_k, U\in(x,x+h)\mid T)\\
&= \sum_{k=2}^n p_{n,k}(b)\left(kh\int_{T_{k-1}}^{T_k}\mu(t)dt + O(h)\right),
\end{align*}
where $p_{n,k}(b) = \binom{n-b-1}{k-2}/\binom{n-1}{k-1}$ is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9).
Marginalizing $T$ gives
\begin{align*}
\mathbbm{P}(C_h) &\simeq h\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right] = h\sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right),
\end{align*}
where $\nu(t)\equiv\int\mu(t)dt$.
% Now, since $\mu(t) > 0$ we have $\nu(t) \nearrow$.
% With $\pi_k(t_k)$ denoting the marginal density of $T_k$, define rv $X_k \equiv \nu(T_k)$, which has density $f_{X_k}(x_k) = \pi_k(\nu^{-1}(x_k))$.
Using eqns 5-8 from \cite{Polanski2003-kg} we can write
\begin{align*}
\mathbbm{E}\left[\nu(T_{k})\right] &= \sum_{j=k}^n A_j^k\int_0^\infty\nu(t)q_j(t)dt
\end{align*}
where
\begin{align*}
A_j^k &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
A_n^n &\equiv 1,\\
q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
\end{align*}
and $\eta(t)$ is the haploid effective population size history.
Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present.
This can be made explicit by defining the inhomogeneous Poisson process intensity function $\lambda_j(t) = \binom{j}{2}/\eta(t)$.

@@

\begin{align*}
\xi_b &= \sum_{k=2}^n p_{n,k}(b) k \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right)\\
&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^{n}B_j^k\int_0^\infty\nu(t)q_j(t)dt
\end{align*}
where
\[
B_j^k \equiv
\begin{cases}
A_{k-1}^{k-1}, j=k-1\\
A_j^k - A_j^{k-1}, k\le j\le n\\
\end{cases}
\]
Exchanging the order of integration and summation yields a manifestly linear expression for the expected SFS as a function of the integrated mutation intensity $\nu(t)$:
\[
\xi_b = \left<\psi_b, \nu\right>,
\]
where
\begin{align*}
\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k q_j(t).
\end{align*}

>>>>

\title{Coalescent inference of mutation spectrum histories from sample frequency spectra}

@@

Let $C_h$ denote the event that there is a mutation with label $U\in(x,x+h)\subseteq(0,1)$ that subtends $b$ haplotypes in the sample, and let $I_k$ denote the event that this mutation arose at a time when the sample had $k$ ancestral lines.
For small $h$ we then have
\begin{align*}
\mathbbm{P}(C_h\mid T) &= \sum_{k=2}^n \mathbbm{P}(C_h, I_k\mid T) = \sum_{k=2}^n p_{n,k}(b)\mathbbm{P}(I_k, U\in(x,x+h)\mid T)\\
&= \sum_{k=2}^n p_{n,k}(b)\left(kh\int_{T_{k-1}}^{T_k}\mu(t)dt + O(h)\right),
\end{align*}
where $p_{n,k}(b) = \binom{n-b-1}{k-2}/\binom{n-1}{k-1}$ is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9).
Marginalizing $T$ gives
\begin{align*}
\mathbbm{P}(C_h) &\simeq h\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right] = h\sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right),
\end{align*}
where $\nu(t)\equiv\int_0^t\mu(t)dt$.
% Now, since $\mu(t) > 0$ we have $\nu(t) \nearrow$.
% With $\pi_k(t_k)$ denoting the marginal density of $T_k$, define rv $X_k \equiv \nu(T_k)$, which has density $f_{X_k}(x_k) = \pi_k(\nu^{-1}(x_k))$.
Using eqns 5-8 from \cite{Polanski2003-kg} we can write
\begin{align*}
\mathbbm{E}\left[\nu(T_{k})\right] &= \sum_{j=k}^n A_j^k\int_0^\infty\nu(t)q_j(t)dt
\end{align*}
where
\begin{align*}
A_j^k &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
A_n^n &\equiv 1,\\
q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
\end{align*}
and $\eta(t)$ is the haploid effective population size history.
Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present.
This can be made explicit by defining the inhomogeneous Poisson process intensity function $\lambda_j(t) = \binom{j}{2}/\eta(t)$.

@@

\begin{align*}
\xi_b &= \sum_{k=2}^n p_{n,k}(b) k \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right)\\
&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^{n}B_j^k\int_0^\infty\nu(t)q_j(t)dt
\end{align*}
where
\[
B_j^k \equiv
\begin{cases}
A_{k-1}^{k-1},\ j=k-1\\
A_j^k - A_j^{k-1},\ k\le j\le n\\
\end{cases}
\]
Using the definition of $\nu$ and exchanging integration order gives
\begin{align*}
\int_0^\infty\nu(t)q_j(t)dt &= \int_0^\infty \left(\int_0^t\mu(x)dx\right) q_j(t)dt\\
&= \int_0^\infty \int_0^\infty 1_{[0<x<t]}(x)\mu(x)dx\ q_j(t)dt\\
&= \int_0^\infty \left(\int_0^\infty 1_{[0<x<t]}(x)q_j(t)dt\right)\mu(x) dx\\
&= \int_0^\infty \left(1-\int_0^x q_j(t)dt\right)\mu(x) dx\\
&= \int_0^\infty \left(1-Q_j(t)\right)\mu(x) dx,
\end{align*}
where $Q_j$ is the distribution function corresponding to density $q_j$.

We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
\[
\xi_b = \left<\psi_b, \mu\right>,
\]
where
\begin{align*}
\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k \left(1-Q_j(t)\right).
\end{align*}
